Bayesian analysis for heterosis detection in RNAseq data

### 2016-04-14 slides/CU/JaradNiemi_CSU.tex
Add CU slides

\documentclass[handout]{beamer}

@@

\title[Bayesian RNAseq heterosis ]{Bayesian analysis for heterosis detection in RNAseq data}
\author[Jarad Niemi]{Dr. Jarad Niemi \\ joint with Eric Mittman, Will Landau, and Dr. Dan Nettleton}
\institute[Iowa State]{Iowa State University}
\date{\today}

\begin{document}

@@

<<chunk_options, echo=FALSE, message=FALSE>>=
opts_chunk$set(fig.width=6, 
               fig.height=5, 
               out.width='.8\\linewidth', 
               fig.align='center', 
               size='tiny',
               echo=FALSE)
library(ggplot2)
@

@@

\item Hierarchical structure:
\begin{columns}
\begin{column}{0.5\linewidth}
\[ \begin{array}{rl}
\phi_g &\stackrel{ind}{\sim} N(\eta_\phi, \sigma_\phi^2) \\
\alpha_g &\stackrel{ind}{\sim} La(\eta_\alpha, \sigma_\alpha) \\
\delta_g &\stackrel{ind}{\sim} La(\eta_\delta, \sigma_\delta) \\
\psi_g &\stackrel{ind}{\sim} N(\eta_\psi, \sigma_\psi^2) 
\end{array} \]
\end{column} \pause
\begin{column}{0.5\linewidth}
<<>>=
source("../../sandbox/Laplace.R")
x = seq(-4,4,by=0.1)
d = data.frame(rbind(data.frame(density="normal", x=x, y=dnorm(x)),
                     data.frame(density="Laplace", x=x, y=dlaplace(x,0,1/sqrt(2)))))
ggplot(d, aes(x=x,y=y,color=density)) + 
  geom_line(size=2) + 
  labs(y="f(x)")
@
\end{column}
\end{columns}
\end{itemize}
\end{frame}

@@

<<echo=TRUE, eval=FALSE>>=
# set up data
# ...
# ...

# get hyperparameter estimates
hyperparameter = get_indep_est(counts.t,group,geneid.t)

# change data to long format
data_stan = w.to.l(counts.t,group)

# compile Stan model (this can take some time, but only needs to be done once)
sg_model = stan_model("sg_model.txt")

# use initial values to obtain hyperparameter estimates
# ...
# ...

# loop over all genes
library(doMC)
registerDoMC()

posterior = dlply(stan_data, .(gene), function(x) {
  dat = list(x, hyperparameter)
  sampling(sg_model, dat, chains=4, iter=2000)
}, .parallel=TRUE)
@

@@

\begin{frame}
\frametitle{}
<<stepping_out_examples>>=
lim = 5
target = function(x) dnorm(x, -2)/2 + dnorm(x, 2)/2
step_out_plot = function(theta, u, w, lim.=lim) {
  curve(target, -lim., lim., axes=F, frame=TRUE, xlab='', ylab='')
  points(theta, u, col='red', pch=19)
  abline(h=u, lty=3, col='gray')

  # Left endpoint
  L = theta - runif(1,0,.5)
  points(L,u)
  while(u<target(L)) points(L <- L-w, u)
  points(L,u, pch=8)

  # Right endpoint
  R = L + w
  points(R,u)
  while(u<target(R)) points(R <- R+w, u)
  points(R,u, pch=8)

  legend('topleft', c('Current', 'Endpoints'), pch=c(19,8), col=c('red','black'))
}

opar = par(mfrow=c(2,1), mar=rep(0.1,4))
w = 1
step_out_plot(2,.15,w)
step_out_plot(2,.05,w)
par(opar)
@
\end{frame}

>>>>

\documentclass[handout]{beamer}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%

\usepackage{framed}
\makeatletter
\newenvironment{kframe}{%
 \def\at@end@of@kframe{}%
 \ifinner\ifhmode%
  \def\at@end@of@kframe{\end{minipage}}%
  \begin{minipage}{\columnwidth}%
 \fi\fi%
 \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
 \colorbox{shadecolor}{##1}\hskip-\fboxsep
     % There is no \\@totalrightmargin, so:
     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
 \MakeFramed {\advance\hsize-\width
   \@totalleftmargin\z@ \linewidth\hsize
   \@setminipage}}%
 {\par\unskip\endMakeFramed%
 \at@end@of@kframe}
\makeatother

\definecolor{shadecolor}{rgb}{.97, .97, .97}
\definecolor{messagecolor}{rgb}{0, 0, 0}
\definecolor{warningcolor}{rgb}{1, 0, 1}
\definecolor{errorcolor}{rgb}{1, 0, 0}
\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

\usepackage{alltt}

@@

\title[Bayesian RNAseq heterosis ]{Bayesian analysis for heterosis detection in RNAseq data}
\author[Jarad Niemi]{Dr. Jarad Niemi \\ joint with Eric Mittman, Will Landau, and Dr. Dan Nettleton}
\institute[Iowa State]{Iowa State University}
\date{\today}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}

@@

@@

\item Hierarchical structure:
\begin{columns}
\begin{column}{0.5\linewidth}
\[ \begin{array}{rl}
\phi_g &\stackrel{ind}{\sim} N(\eta_\phi, \sigma_\phi^2) \\
\alpha_g &\stackrel{ind}{\sim} La(\eta_\alpha, \sigma_\alpha) \\
\delta_g &\stackrel{ind}{\sim} La(\eta_\delta, \sigma_\delta) \\
\psi_g &\stackrel{ind}{\sim} N(\eta_\psi, \sigma_\psi^2) 
\end{array} \]
\end{column} \pause
\begin{column}{0.5\linewidth}
\begin{knitrout}\tiny
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}

{\centering \includegraphics[width=.8\linewidth]{figure/unnamed-chunk-1-1}

}

\end{knitrout}
\end{column}
\end{columns}
\end{itemize}
\end{frame}

@@

\begin{knitrout}\tiny
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# set up data}
\hlcom{# ...}
\hlcom{# ...}

\hlcom{# get hyperparameter estimates}
\hlstd{hyperparameter} \hlkwb{=} \hlkwd{get_indep_est}\hlstd{(counts.t,group,geneid.t)}

\hlcom{# change data to long format}
\hlstd{data_stan} \hlkwb{=} \hlkwd{w.to.l}\hlstd{(counts.t,group)}

\hlcom{# compile Stan model (this can take some time, but only needs to be done once)}
\hlstd{sg_model} \hlkwb{=} \hlkwd{stan_model}\hlstd{(}\hlstr{"sg_model.txt"}\hlstd{)}

\hlcom{# use initial values to obtain hyperparameter estimates}
\hlcom{# ...}
\hlcom{# ...}

\hlcom{# loop over all genes}
\hlkwd{library}\hlstd{(doMC)}
\hlkwd{registerDoMC}\hlstd{()}

\hlstd{posterior} \hlkwb{=} \hlkwd{dlply}\hlstd{(stan_data,} \hlkwd{.}\hlstd{(gene),} \hlkwa{function}\hlstd{(}\hlkwc{x}\hlstd{) \{}
  \hlstd{dat} \hlkwb{=} \hlkwd{list}\hlstd{(x, hyperparameter)}
  \hlkwd{sampling}\hlstd{(sg_model, dat,} \hlkwc{chains}\hlstd{=}\hlnum{4}\hlstd{,} \hlkwc{iter}\hlstd{=}\hlnum{2000}\hlstd{)}
\hlstd{\},} \hlkwc{.parallel}\hlstd{=}\hlnum{TRUE}\hlstd{)}
\end{alltt}
\end{kframe}
\end{knitrout}

@@

\begin{frame}
\frametitle{}
\begin{knitrout}\tiny
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}

{\centering \includegraphics[width=.8\linewidth]{figure/stepping_out_examples-1}

}

\end{knitrout}
\end{frame}
